\section{Results from Prior NSF Support}

This is PI Pandey's first NSF proposal.

For PI Owens, the NSF award whose outcomes most influenced this grant was the recently completed NSF Algorithms in the Field award CCF-1637442, ``AitF\@: Collaborative Research: Theory and Implementation of Dynamic Data Structures for the GPU'', with PI Farach-Colton. This work focused on the design and implementation of GPU-friendly, dynamic, general-purpose data structures.
\textbf{Intellectual Merit.} In this AitF award, the PIs designed and built numerous dynamic data structures that marked their first implementation on GPUs: linked lists and hash tables~\cite{Ashkiani:2018:ADH}, log-structured merge trees~\cite{Ashkiani:2018:GLA}, B-trees~\cite{Awad:2019:EAH}, quotient filters~\cite{Geil:2018:QFA}, and dynamic graphs~\cite{Awad:2020:DGO}. RXMesh's mesh data structure~\cite{Mahmoud:2021:RAG} was a result of this award. As well, the PIs recently designed and implemented the fastest GPU static hash tables~\cite{Awad:2023:AAI}.

\textbf{Broader Impacts.} Each of the abovementioned projects is available as open-source software, a significant reason why this work is highly cited. The dynamic graphs were integrated into Gunrock~\cite{Awad:2020:DGO,Wang:2017:GGG}, which subsequently was selected as the performance reference for the DARPA HIVE program and in turn integrated into NVIDIA's open-source RAPIDS framework for data science. Finally, NVIDIA's cuCollections library currently only supports static and dynamic maps but these were both influenced by the PIs' work and we expect as this library grows, it will incorporate the abovementioned work either directly or as the basis for NVIDIA's implementation and improvements.

PIs Bender and Farach-Colton's most relevant NSF-funded project ``CCF-BSF\@: AF\@: Small: Collaborative Research: The Dictionary Problem Considered''
(CCF-BSF 1716252, \$250,000, and CCF-BSF 1716252, \$250,000, 9/1/17--8/31/20).
The project is investigating algorithms inspired by systems requirements in dictionary data structures.
%\rob{The proposed research will investigate performance and functionality limitations of mmap that prevent its use in big data applications.}
\textbf{Intellectual Merit.} This grant has funded the following
publications~\cite{AgrawalBeDa20,AgrawalBeFi20,%
ArkinDaGa20,Ashkiani:2018:ADH,Ashkiani:2018:GLA,BenderChDa20,BenderCoFa19,BenderDaJo20,BenderFaGo18,%
BenderFaKu19,BenderGoMe20,BenderKoKu20,ChenMcSi18,ConwayBaJi17b,ConwayFaSh18,ConwayKnJi19,%
DasAgBe20,DasTsDu18,DasTsDu19,BerceaEv20a,BerceaEv20b,%
ZhanCoJi18,ZhanJaPo18,ZhanJaPo18,Mayer18,Pandey19,PandeyABFJP18Cell,PandeyBJP17,%
PandeyBJP17a,PandeyBJP17b,Singh18,SinghMaBe20,%
JavanmardGaDa19DISC,%
GoswamiMeMe18,%
Geil:2018:QFA,pandey2020timely,%
EvenMeRa18,ConwayFaSh18,Awad:2019:EAH,BenderFiGi19,Farach-ColtonLiTs18,BenderKoPe18}.

\textbf{Broader Impact.} The project has produced several software artifacts and involved several Ph.D.\ students, including two woman and a Hispanic student, as well as a woman undergraduate student.
